---
title: "Schwarz-Christoffel Conformal Mappings"
author: "Peter DeWitt"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: true
    depth: 2
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
library(sccm)
knitr::opts_chunk$set(collapse = TRUE)
```

The `sccm` package provides some tools for working with two-dimensional
polygons, especially convex hulls, and mapping the data within these
polygons/convex hulls to other polygons via Schwarz-Christoffel conformal
mappings.

*Acknowledgements:*  I would be remiss to not acknowledge two other R packages
which greatly influenced the design of the `sccm` package.

* [`conformal`: https://r-forge.r-project.org/R/?group_id=973)](https://r-forge.r-project.org/R/?group_id=973) by Nick Ellis
* [`contoureR`: https://cran.r-project.org/web/packages/contoureR/index.html](https://cran.r-project.org/web/packages/contoureR/index.html) by
  Nicholas Hamilton.

The Schwarz-Christoffel conformal mappings are well described in 

* Trefethen, Lloyd N. "Analysis and design of polygonal resistors by conformal mapping." Zeitschrift für angewandte Mathematik und Physik ZAMP 35.5 (1984): 692-704.

Corresponding FORTRAN libraries are vailable from 
[SCPACK: http://www.netlib.org/conformal/](http://www.netlib.org/conformal/).

The particular set of functions provided by the `sccm` package are a minimum set
of tools needed to map 2D data to a pre-defined support for regression models.
The details and examples of the regression models are part of Peter DeWitt's
Ph.D. dissertation.  More details on this can be found in the (soon to be
released `cpr` package.

```{r}
set.seed(42)

# Required packages to re-create the following examples
library(sccm)
```

# Coordinate Translation
There are two functions provided to make it easy to move between Cartesian and
polar coordinates:

1. `polar2cartesian(r, theta)` and
2. `cartesian2polar(x, y)`.

These function are based on the relationship
$$(x, y) = (r \cos(\theta), r \sin(\theta)).$$

Both functions expect that the inputs are numeric vectors of equal length or, if
one input is length 1, it will be repeated to match the length of the other
input.  Examples below

```{r}
# return the Cartesian coordinates of the point on the unit circle with and
# angle of 2 pi/ 3
cart <- polar2cartesian(r = 1, theta = 2 * pi / 3)
cart
all.equal(cart, matrix(c(-0.5, sqrt(3) / 2), nrow = 1)
)

# Return a disk of a given radius, or a radii of a
disk3 <- polar2cartesian(r = 3, theta = seq(0, 2 * pi, length = 500L))
disk4 <- polar2cartesian(r = 4, theta = seq(0, 2 * pi, length = 500L))
radii <- polar2cartesian(r = seq(0, 4, length = 100L), theta = 0.75 * pi)

plot(disk4, col = "black", asp = 1, xlab = "", ylab = "")
points(disk3, col = "red")
points(radii, col = "blue")
```

`cartesian2polar` translates an $(x, y)$ Cartesian coordinates to $(r, \theta)$
polar coordinates.

```{r}
cartesian2polar(-0.5, sqrt(3)/2)
all.equal(cartesian2polar(-0.5, sqrt(3)/2), 
          matrix(c(1, 2 * pi / 3), nrow = 1))
```

# Polygons and Convex Hulls
The conformal mappings will require the use of polygons.  The functions
`sccm::polygon` and `sccm::convex_hull` are provided to make defining the
polygons easy.  The function `is_in` determines if a given point is either
inside, on an edge, or outside of a given polygon.

## `polygon`
The expected input for `polygon` are two numeric vectors, or a `matrix`, or a
`data.frame`.  Any R object that can be coerced into a two column `matrix` could
be passed to `sccm::polygon`.

For the set of vertices 
$\left\{ (1, -1), (1, 0), (2, 2), (3, -2) \right\}$ the polygon is defined
via:

```{r}
pg <- polygon(x = c( 1, 1, 2,  3), 
              y = c(-1, 0, 2, -2))

pg

class(pg)

str(pg)

plot(pg)
```

`pg` is a `sccm_pg` object.  This object returns the vertices entered, in order
as entered, as a $n \times 2$ matrix.  The `beta` element are the exterior
angles at each vertex.  These exterior angles are determine by the `atan2`
function.
`atan2(y, x)` is similar to `atan(y / x)` only that the former considers the
sign of both $y$ and $x$ to determine the quadrant and return the angel from
origin between 0 and $2\pi,$  whereas the later returns an angle between
$-\pi/2$ and $\pi/2.$

$$\beta_{j} = \text{atan2} \left( \overrightarrow{v_{j-1}v_{j}} \times \overrightarrow{v_{j}v_{j+1}}, \overrightarrow{v_{j-1}v_{j}} \cdot \overrightarrow{v_{j}v_{j+1}}\right)$$

The cross product of the vectors by the dot product provides the needed inputs
for determining the exterior angle of each vertex.

It is important to note that `sccm::polygon` returns the vertices of the polygon
in the same order as they inputted.  For the conformal mappings we will need the
vertices to be entered in anti-clockwise order.  The user is expected to
determine the correct sequence when constructing a polygon.  

To illustrate the ordering of the vertices we will construct a five pointed star
polygon.

```{r}
star <- 
  rbind(polar2cartesian(1.0, seq(0, 1.6, by = 0.4) * pi),
        polar2cartesian(0.5, seq(0.2, 1.8, by = 0.4) * pi)) 

plot(polygon(star))
```

Notice that the vertices on the unit disk are listed before the vertices on a
disk of radius 0.5 and the plot reflex this.  While the plot is a polygon,
albeit an irregular polygon, it does not look like a star, a regular concave
polygon.  Next we will put the vertices in the correct order and generate a new
plot.  This `star_pg` polygon will be used in several examples to follow.

```{r}
star <- star[rep(1:5, each = 2) + rep(c(0, 5), times = 5), ]
star_pg <- polygon(star)
plot(star_pg)
```

## `convex_hull`
A convex hull is the smallest convex envelop containing all the points of a set.
The construction of these convex hulls is done via Andrew's Monotone Theorem, a
$O(n log n)$ order algorithm.  A brute force approach for finding convex hulls
can be $O(n^3).$

The function `convex_hull` will accept a $n \times 2$ `matrix` or `data.frame`
or two numeric vectors given the $(x, y)$ coordinates of the data.  The function
will then return the vertices of the convex hull along with the exterior angles
at each vertices and the row index of the original data for each of the
vertices.  It is also worth noting that the origin data is returned within the
`list` structure of `sccm_ch` objects.

```{r}
xvec = runif(150)
yvec = 2 * xvec + rnorm(150)

ch <- convex_hull(xvec, yvec)

ch

class(ch)

str(ch)

plot(ch)
```

Using the `mtcars` data set, we can build a convex hull about the weight of
vertices and the miles per gallon.

```{r}
plot(convex_hull(mtcars[, c("mpg", "wt")]))
title(xlab = "MT Cars: MPG", ylab = "MT Cars: Weight")
```

## `is_in`


# Schwarz-Christoffel Conformal Mappings

## Disk to Polygon

## Polygon to Disk

## Polygon to Polygon
