---
title: "Schwarz-Christoffel Conformal Mappings"
author: "Peter DeWitt"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: true
    depth: 2
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
library(sccm)
knitr::opts_chunk$set(collapse = TRUE)
```

The `sccm` package provides some tools for working with two-dimensional
polygons, especially convex hulls, and mapping the data within these
polygons/convex hulls to other polygons via Schwarz-Christoffel conformal
mappings.

*Acknowledgements:*  I would be remiss to not acknowledge two other R packages
which greatly influenced the design of the `sccm` package.

* [`conformal`: https://r-forge.r-project.org/R/?group_id=973)](https://r-forge.r-project.org/R/?group_id=973) by Nick Ellis
* [`contoureR`: https://cran.r-project.org/web/packages/contoureR/index.html](https://cran.r-project.org/web/packages/contoureR/index.html) by
  Nicholas Hamilton.

The Schwarz-Christoffel conformal mappings are well described in 

* Trefethen, Lloyd N. "Analysis and design of polygonal resistors by conformal mapping." Zeitschrift für angewandte Mathematik und Physik ZAMP 35.5 (1984): 692-704.

Corresponding FORTRAN libraries are vailable from 
[SCPACK: http://www.netlib.org/conformal/](http://www.netlib.org/conformal/).

The particular set of functions provided by the `sccm` package are a minimum set
of tools needed to map 2D data to a pre-defined support for regression models.
The details and examples of the regression models are part of Peter DeWitt's
Ph.D. dissertation.  More details on this can be found in the (soon to be
released `cpr` package.

```{r}
set.seed(42)

# Required packages to re-create the following examples
library(sccm)
```

# Data Sets
There are two data sets provided in the `sccm` package.  These data sets are
provided to help illustrate the conformal mappings.

### `CircleLimitI`
A circular drawing by M.C. Escher, has been provided as a `data.frame`.  See
`?CircleLimitI` for more information.
```{r}
head(CircleLimitI)
```

```{r, fig.width = 4, fig.height = 4}
plot(CircleLimitI[, c("x", "y")], 
     asp = 1,
     col = cut(CircleLimitI$value, breaks = 2), 
     pch = ".", cex = 3) 
```

### `HexagonalFish`
Another M.C. Escher drawing, useful for showing the mapping between a polygon
and the unit disk or another polygon. See `?HexagonalFish` for more details.

```{r, fig.width = 4, fig.height = 4}
plot(HexagonalFish[, c("x", "y")], 
     asp = 1,
     col = cut(HexagonalFish$value, breaks = 2), 
     pch = ".", cex = 3) 
```


# Coordinate Translation
There are two functions provided to make it easy to move between Cartesian and
polar coordinates:

1. `polar2cartesian(r, theta)` and
2. `cartesian2polar(x, y)`.

These function are based on the relationship
$$(x, y) = (r \cos(\theta), r \sin(\theta)).$$

Both functions expect that the inputs are numeric vectors of equal length or, if
one input is length 1, it will be repeated to match the length of the other
input.  Examples below

```{r}
# return the Cartesian coordinates of the point on the unit circle with and
# angle of 2 pi/ 3
cart <- polar2cartesian(r = 1, theta = 2 * pi / 3)
cart
all.equal(cart, matrix(c(-0.5, sqrt(3) / 2), nrow = 1)
)

# Return a disk of a given radius, or a radii of a
disk3 <- polar2cartesian(r = 3, theta = seq(0, 2 * pi, length = 500L))
disk4 <- polar2cartesian(r = 4, theta = seq(0, 2 * pi, length = 500L))
radii <- polar2cartesian(r = seq(0, 4, length = 100L), theta = 0.75 * pi)

plot(disk4, col = "black", asp = 1, xlab = "", ylab = "")
points(disk3, col = "red")
points(radii, col = "blue")
```

`cartesian2polar` translates an $(x, y)$ Cartesian coordinates to $(r, \theta)$
polar coordinates.

```{r}
cartesian2polar(-0.5, sqrt(3)/2)
all.equal(cartesian2polar(-0.5, sqrt(3)/2), 
          matrix(c(1, 2 * pi / 3), nrow = 1))
```

# Polygons and Convex Hulls
The conformal mappings will require the use of polygons.  The functions
`sccm::polygon` and `sccm::convex_hull` are provided to make defining the
polygons easy.  The function `is_in` determines if a given point is either
inside, on an edge, or outside of a given polygon.

## `polygon`
The expected input for `polygon` are two numeric vectors, or a `matrix`, or a
`data.frame`.  Any R object that can be coerced into a two column `matrix` could
be passed to `sccm::polygon`.

For the set of vertices 
$\left\{ (1, -1), (1, 0), (2, 2), (3, -2) \right\}$ the polygon is defined
via:

```{r}
pg <- polygon(x = c( 1, 1, 2,  3), 
              y = c(-1, 0, 2, -2))

pg

class(pg)

str(pg)

plot(pg)
```

`pg` is a `sccm_pg` object.  This object returns the vertices entered, in order
as entered, as a $n \times 2$ matrix.  The `beta` element are the exterior
angles at each vertex.  These exterior angles are determine by the `atan2`
function.
`atan2(y, x)` is similar to `atan(y / x)` only that the former considers the
sign of both $y$ and $x$ to determine the quadrant and return the angel from
origin between 0 and $2\pi,$  whereas the later returns an angle between
$-\pi/2$ and $\pi/2.$

$$\beta_{j} = \text{atan2} \left( \overrightarrow{v_{j-1}v_{j}} \times \overrightarrow{v_{j}v_{j+1}}, \overrightarrow{v_{j-1}v_{j}} \cdot \overrightarrow{v_{j}v_{j+1}}\right)$$

The cross product of the vectors by the dot product provides the needed inputs
for determining the exterior angle of each vertex.

It is important to note that `sccm::polygon` returns the vertices of the polygon
in the same order as they inputted.  For the conformal mappings we will need the
vertices to be entered in anti-clockwise order.  The user is expected to
determine the correct sequence when constructing a polygon.  

To illustrate the ordering of the vertices we will construct a five pointed star
polygon.

```{r}
star <- 
  rbind(polar2cartesian(1.0, seq(0, 1.6, by = 0.4) * pi),
        polar2cartesian(0.5, seq(0.2, 1.8, by = 0.4) * pi)) 

plot(polygon(star))
```

Notice that the vertices on the unit disk are listed before the vertices on a
disk of radius 0.5 and the plot reflex this.  While the plot is a polygon,
albeit an irregular polygon, it does not look like a star, a regular concave
polygon.  Next we will put the vertices in the correct order and generate a new
plot.  This `star_pg` polygon will be used in several examples to follow.

```{r}
star <- star[rep(1:5, each = 2) + rep(c(0, 5), times = 5), ]
star_pg <- polygon(star)
plot(star_pg)
```

## `convex_hull`
A convex hull is the smallest convex envelop containing all the points of a set.
The construction of these convex hulls is done via Andrew's Monotone Theorem, a
$O(n log n)$ order algorithm.  A brute force approach for finding convex hulls
can be $O(n^3).$

The function `convex_hull` will accept a $n \times 2$ `matrix` or `data.frame`
or two numeric vectors given the $(x, y)$ coordinates of the data.  The function
will then return the vertices of the convex hull along with the exterior angles
at each vertices and the row index of the original data for each of the
vertices.  It is also worth noting that the origin data is returned within the
`list` structure of `sccm_ch` objects.

```{r}
xvec = runif(150)
yvec = 2 * xvec + rnorm(150)

ch <- convex_hull(xvec, yvec)

ch

class(ch)

str(ch)

plot(ch)
```

Using the `mtcars` data set, we can build a convex hull about the weight of
vertices and the miles per gallon.

```{r}
plot(convex_hull(mtcars[, c("mpg", "wt")]))
title(xlab = "MT Cars: MPG", ylab = "MT Cars: Weight")
```

## `is_convex_hull`
This function tests if a polygon is a convex hull or not.
```{r}
is_convex_hull(ch)
is_convex_hull(polygon(c(-1, -1, 1, 1), c(-1, 1, 1, -1)))  # clockwise square
is_convex_hull(polygon(c(-1, -1, 1, 1), c(1, -1, -1, 1)))  # anti-clockwise square
is_convex_hull(star_pg)
```

## `is_anticlockwise`
This function tests if a polygon is a convex hull and if so, are the vertices
listed in anti-clockwise order.  Note: by construction, `sccm_ch` objects have
the vertices listed in anti-clockwise order.

```{r}
is_anticlockwise(ch)
is_anticlockwise(polygon(c(-1, -1, 1, 1), c(-1, 1, 1, -1)))  # clockwise square
is_anticlockwise(polygon(c(-1, -1, 1, 1), c(1, -1, -1, 1)))  # anti-clockwise square
is_anticlockwise(star_pg)
```

## `is_in`

The `is_in` function tests if a point is located on the inside, edge, or
outside of a polygon.  The implementation is based on the winding number
algorithm.

```{r}
# a point on an edge
xx <- 0.75
yy <- (diff(star_pg$vertices[1:2, 2]) / diff(star_pg$vertices[1:2, 1])) * (xx - star_pg$vertices[1, 1]) + star_pg$vertices[1, 2]

is_in(-0.4, 0.1, star_pg) # inside
is_in(0.5, -0.6, star_pg) # outside
is_in(0.5, yy, star_pg)   # on an edge

# send a set of points to is_in
xcoords <- c(-0.4, 0.5, 0.5)
ycoords <- c(0.1, -0.6, yy)
is_in(xcoords, ycoords, star_pg)
summary(is_in(xcoords, ycoords, star_pg))

plot(star_pg, asp = 1)
points(-0.4, 0.1, pch = 16, col = "black") # inside
points(0.5, -0.6, pch = 16, col = "blue")  # outside
points(xx, yy,    pch = 16, col = "red")   # on an edge 
```

# Schwarz-Christoffel Conformal Mappings

The Schwarz-Christoffel conformal mapping (SCCM) from the unit disk to a polygon
are used by three functions:

1. `d2p` for disk to polygon,
2. `p2p` for polygon to disk (an inverse mapping), and
3. `p2p` form one polygon to another via an (inverse) mapping from a polygon to
   the unit disk and a mapping from the disk to the other polygon.

Good references for mathemtatics and the FORTRAN program can be found at:

* Trefethen, Lloyd N. "Numerical computation of the Schwarz-Christoffel transformation." SIAM Journal on Scientific and Statistical Computing 1.1 (1980): 82-102.

* Trefethen, Lloyd N. "Analysis and design of polygonal resistors by conformal mapping." Zeitschrift für angewandte Mathematik und Physik ZAMP 35.5 (1984): 692-704.

* [SCPACK](http://www.netlib.org/conformal/)


## Disk to Polygon

Evaluation of the forward map $w(z)$ for the point $z$ in the unit disk or on
the circumference thereof, requires the computation of the integral

$$w = w(z_0) + C \int\limits_{z_0}^{z} \prod\limits_{j=1}^{N} \left( 1 - \frac{\zeta}{z_j} \right)^{-\beta_j} \mathrm{d}\zeta.$$

The endpoint $z_0$ may be any point in the closed disk where $w(z_0)$ is known
and finite.  Additional detain is in Trefethen (1980).

The function `d2p` provides the mapping of `.data` from the unit disk to a given
polygon, `pg`.  By default, the function will map the unit disk to a square with
vertices at (-1, 1), (1, -1), (1, 1), and (-1, 1).

```{r}
args(d2p)
```

To illustrate this mapping we'll use the `CircleLimitI` data set.

The mapping object will contain the `mapped` data, the `polygon` the data was
mapped into, the original `data` provided

```{r}
circle_to_square <- d2p(CircleLimitI[, c("x", "y")])
str(circle_to_square, max.level = 1L)
```

Plotting the results is simple:
```{r, fig.width = 9, fig.height = 6}
plot(circle_to_square,
     col = cut(CircleLimitI$value, breaks = 2), 
     pch = ".", cex = 3) 
```

Mapping to any simple polygon is possible:
```{r, fig.width = 9, fig.height = 6}
plot(d2p(CircleLimitI[, c("x", "y")], pg = star_pg),
     col = cut(CircleLimitI$value, breaks = 2), 
     pch = ".", cex = 3) 
```

Mapping the unit disk to the square and star shapes viewed with color disks:
```{r, fig.width = 9, fig.height = 6}
color_disk <- 
  Reduce(rbind,
    c(
      mapply(function(r, theta) {cbind(sccm::polar2cartesian(r, theta), r)},
           r = seq(0.1, 0.9, by = 0.1), 
           MoreArgs = list(theta = seq(0, 2 * pi, length = 100L)),
           SIMPLIFY = FALSE),
      mapply(function(r, theta) {cbind(sccm::polar2cartesian(r, theta), r)},
             theta = seq(0, 2, by = 0.25) * pi,
             MoreArgs = list(r = seq(0, 0.99, length = 20L)),
             SIMPLIFY = FALSE)
      ))

color_disk_squared <- sccm::d2p(color_disk[, 1:2])
color_disk_stared  <- sccm::d2p(color_disk[, 1:2],
                                pg = sccm::polygon(star))

plot(color_disk_squared, col = cut(color_disk[, 3], breaks = 6), pch = ".", cex = 3)
plot(color_disk_stared,  col = cut(color_disk[, 3], breaks = 6), pch = ".", cex = 3)
```


## Polygon to Disk
An inverse mapping taking two arguments, `.data` to map and the polygon
containing the `.data.`
```{r, fig.width = 6, fig.height = 4}
plot(p2d(HexagonalFish[, c("x", "y")]), 
     col = cut(HexagonalFish$value, breaks = 2), 
     pch = ".", cex = 3)
```

## Polygon to Polygon
To map `.data` within polygon `pg1` to another polygon `pg2` use the function
`p2p`.  Default settings will map `.data` within its convex hull to the square
with vertices at (-1, 1), (1, -1), (1, 1), and (-1, 1).
```{r}
args(p2p)
```

Some examples:
```{r, fig.width = 9, fig.height = 6}
square_fish <- p2p(HexagonalFish[, 1:2])
plot(square_fish, col = cut(HexagonalFish[, 3], breaks = 2))
```

```{r, fig.width = 9, fig.height = 6}
star_fish <- p2p(HexagonalFish[, 1:2], pg2 = star_pg)
plot(star_fish, col = cut(HexagonalFish[, 3], breaks = 2))
```
